---
title: "Regression Models: Course Project"
author: "Amit Kohli"
date: "February 15, 2016"
output: pdf_document
fontsize: 10pt
geometry: margin=.5in
---

```{r prepare_environment, echo=FALSE, include=FALSE}
setwd("~/datasciencecoursera/Regression Models/Course Project")
library(ggplot2)
library(knitr)
library(datasets)
library(xtable)
library(dplyr)
data(mtcars)
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$am <- factor(mtcars$am)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
attach(mtcars)
opts_chunk$set(results = 'hold',fig.path = 'figure/')
```
  
**Executive Summary**  This report analyzes mtcars dataset to explore if an automatic or manual transmission is better for cars in terms of miles per gallon (mpg) and quantifying the difference between both types of transmissions. The reports uncovers a strong relationship between vehicle weight and transmission type, but there is still a statistically significant increase in the mean value of gas mileage attributable to manual transmissions -- *between 0.05 and 5.8 mpg.*

**Exploratory Data Analysis** Dataset has 32 observations on 11 variables.  
```{r, echo=FALSE}
dim(mtcars)
head(mtcars)
tail(mtcars)
```   
By checking correlations, weight variable affects the fuel economy the most - and by plotting mpg and weight against transmission types (*Figures 1 & 2*) -- the report reveals that weight difference for transmissions is stronger than mpg difference.  

**Inference** The report makes the null hypothesis as the MPG of the automatic and manual transmissions are from the same population by utilizing the the two sample T-test function. Since the p-value is 0.00137, we reject our null hypothesis. So, the automatic and manual transmissions are from different populations. And the mean for MPG of manual transmitted cars is about 7 more than that of automatic transmitted cars.
```{r, echo=FALSE}
result <- t.test(mtcars[mtcars$am == 0, ]$mpg, mtcars[mtcars$am == 1, ]$mpg)
result$p.value
result$estimate
```  
  
**Regression Analysis** The report runs several models to understand relationship between different variables, starting with the full model below. As the p-value is much less than 0.05, we reject the null hypothesis. Hence there is a significant relationship between the variables in the linear regression model of the dataset faithful. Results of __Model 1__:
```{r, echo=FALSE}
model1 <- lm(mpg ~ ., data=mtcars)
summary(model1)$call
paste("Residual Standard Error: ",summary(model1)$sigma)
paste("Degrees of Freedom: ",summary(model1)$df[2])
paste("Adjusted R-squared: ",summary(model1)$adj.r.squared , " (model can explain 78% of the variance of the MPG variable)")
model1.pvalue<-summary(model1)$fstatistic
model1.pvalue<-pf(model1.pvalue[1],model1.pvalue[2],model1.pvalue[3],lower.tail=F)
attributes(model1.pvalue)<-NULL
paste("p-value: ",model1.pvalue)
```  

Next, the report uses backward selection to select some statistically significant variables. Results of __Model 2__:
```{r, echo=FALSE,results='hide'}
model2 <- step(model1, k=log(nrow(mtcars)))
```
```{r, echo=FALSE}
summary(model2)$call
paste("Residual Standard Error: ",summary(model2)$sigma)
paste("Degrees of Freedom: ",summary(model2)$df[2])
paste("Adjusted R-squared: ",summary(model2)$adj.r.squared, " (model can explain 83% of the variance of the MPG variable)")
model2.pvalue<-summary(model2)$fstatistic
model2.pvalue<-pf(model2.pvalue[1],model2.pvalue[2],model2.pvalue[3],lower.tail=F)
attributes(model2.pvalue)<-NULL
paste("p-value: ",model2.pvalue)
```  

Next, the report fits the simple model with MPG as the outcome variable and Transmission as the predictor variable. It shows that on average, a car has 17.147 mpg with automatic transmission, and if it is manual transmission, 7.245 mpg is increased. Results of __Model 3__:
```{r, echo=FALSE}
model3<-lm(mpg ~ am, data=mtcars)
summary(model3)$call
#summary(model3)$coef
paste("Residual Standard Error: ",summary(model3)$sigma)
paste("Degrees of Freedom: ",summary(model3)$df[2])
paste("Adjusted R-squared: ",summary(model3)$adj.r.squared, " (model can explain 34% of the variance of the MPG variable)")
model3.pvalue<-summary(model3)$fstatistic
model3.pvalue<-pf(model3.pvalue[1],model3.pvalue[2],model3.pvalue[3],lower.tail=F)
attributes(model3.pvalue)<-NULL
paste("p-value: ",model3.pvalue)
```  

Lastly, according to the scatter plot (*Figure 3*), it indicates that there appears to be an interaction between "wt" and "am" variables, since automatic cars tend to weigh heavier than manual cars. Results of __Model 4__:
```{r, echo=FALSE}
model4<-lm(mpg ~ wt + qsec + am + wt:am, data=mtcars)
summary(model4)
#summary(model4)$call
#paste("Residual Standard Error: ",summary(model4)$sigma)
#paste("Degrees of Freedom: ",summary(model4)$df[2])
#paste("Adjusted R-squared: ",summary(model4)$adj.r.squared, " (model can explain 88% of the variance of the MPG variable)")
#model4.pvalue<-summary(model4)$fstatistic
#model4.pvalue<-pf(model4.pvalue[1],model4.pvalue[2],model4.pvalue[3],lower.tail=F)
#attributes(model4.pvalue)<-NULL
#paste("p-value: ",model4.pvalue)
```  
The report selects the final model (*Model 4*) with the highest Adjusted R-squared value (88%), "mpg ~ wt + qsec + am + wt:am".  Model 4 results show that when "wt" (weight lb/1000) and "qsec" (1/4 mile time) remain constant, cars with manual transmission add 14.079 + (-4.141)*wt more MPG (miles per gallon) on average than cars with automatic transmission. That is, a manual transmitted car that weighs 2000 lbs have 5.797 more MPG than an automatic transmitted car that has both the same weight and 1/4 mile time.
```{r, echo=FALSE}
#summary(model4)$coef
```  

**Residual Analysis and Diagnostics** According to the residual plots (*Figure 4*), we can verify the following underlying assumptions:  
1. The Residuals vs. Fitted plot shows no consistent pattern, supporting the accuracy of the independence assumption.  
2. The Normal Q-Q plot indicates that the residuals are normally distributed because the points lie closely to the line.  
3. The Scale-Location plot confirms the constant variance assumption, as the points are randomly distributed.  
4. The Residuals vs. Leverage argues that no outliers are present, as all values fall well within the 0.5 bands.  

As for the Standardized Difference of the Beta, the measure of how much an observation has effected the estimate of a regression coefficient.  Results below prove that the report meets all basic assumptions of linear regression model.  
```{r, echo=FALSE}
paste("dfbetas: ",sum((abs(dfbetas(model4)))>1))
```  

\pagebreak
\begin{center}
Appendix
\end{center}

## Figure 1 - Boxplot of MPG vs. Transmission  
```{r, echo=FALSE}
par(mfrow = c(1, 2))
transFactor <- factor(mtcars$am, levels=0:1, labels=c("Automatic", "Manual"))
plot(transFactor, mtcars$mpg, xlab="Transmission Type", ylab="MPG",col="green")
plot(transFactor, mtcars$wt, xlab="Transmission Type", ylab="Weight (1000lb)",col="blue")
```  

## Figure 2 - Pair Graph of Motor Trend Dataset 
```{r, echo=FALSE}
pairs(mtcars, panel=panel.smooth, col=blues9)
```  

## Figure 3 - Scatter Plot of MPG vs. Weight by Transmission  
```{r, echo=FALSE}
ggplot(mtcars, aes(x=wt, y=mpg, group=am, color=am, height=3, width=3)) + geom_point() +  
scale_color_discrete(labels=c("Automatic", "Manual")) + 
xlab("weight")
```

## Figure 3(a) - Scatter Plot of MPG vs. Cylinder
```{r, echo=FALSE}
ggplot(mtcars, aes(x=cyl, y=mpg, group=cyl, color=cyl, height=3, width=3)) + geom_point() + scale_color_discrete(labels=c("4", "6","8")) + 
xlab("Cylinder")
```  

## Figure 4 - Residual Plots
```{r,echo=FALSE,fig.height=12, fig.width=10}
par(mfrow = c(2, 2))
plot(model4)
```  


<td>[GitHub Repo](https://github.com/akohli70/Regression-Models/blob/master/motor_trend.Rmd)</td>